---
title: "Natural Language Processing Methods are Sensitive to Sub-Clinical Linguistic Differences in Schizophrenia Spectrum Disorders"
author: "Sunny X. Tang, Reno Kriz, Sunghye Cho, Suh Jung Park, Jenna Harowitz, Raquel E. Gur, Mahendra T. Bhati, Daniel H. Wolf, João Sedoc, Mark Y. Liberman"
date: "1/14/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

load("datshort.R")
load("datlong.R")
load("justBERTdif.R")
load("justBERTdif_rmout.R")

library(effsize)
library(nlme)
library(ggplot2)
library(Hmisc)
library(plyr)
library(arsenal)
library(lme4)
library(dplyr)
```

## Summary
- Purpose: this code summarizes basic sample statistics and investigates group effect on BERT next sentence prediction as well as divergence from interviewer prompt.

## Data Source
- Cohort 1 "Worden" - includes control, schizophrenia and schizoaffective disorder
- Cohort 2 "Wolf Pilot" - includes control and schizophrenia participants; combines data from both positive and neutral stimuli
- Final analysis: DID NOT DROP any length sentences. Previous analyses examined the effect of excluding short sentences with fewer than 2 or fewer than 5 words with the thought that longer sentences would be more illustrative of differences in coherence. This did not produce a difference in the results, so these were not included in the final analysis shown here.
- BERT Sentence pairs include both participant-participant and interviewer-participant pairs
- TLC ratings (missing for n=3 due to recording not available)
- Withing subject normalized weighted scores were calculated: weightedbert = (l1b1 + l2b2 + l3b3 + ... )/(l1 + l2 + l3 + ...), but were not reported in the manuscript because this did not affect the results.

## Calculation of summary variables 
```{r, include=FALSE}
######## Calculating mean BERT scores by subject ##########
justsubjectbert <- cbind(rep(0,31), rep(0,31), rep(0,31), rep(0,31), rep(0,31)) %>% as.data.frame()
colnames(justsubjectbert) <- c("bblid", "meanbert", "numbersentences", "meansentlength", "meanweightbert")

for (i in 1:31) {
      temp1 <- filter(datlong, bblid==datshort$bblid[i])
      temp1$bblid <- as.character(temp1$bblid)
      justsubjectbert[i,1] <- temp1[i,1]
      justsubjectbert[i,2] <- sum(temp1$bert)/dim(temp1)[1]
      justsubjectbert[i,3] <- dim(temp1)[1]
      justsubjectbert[i,4] <- sum(temp1$s2length)/dim(temp1)[1]
      justsubjectbert[i,5] <- sum(temp1$weightbert)/sum(temp1$s2length)
}

datshort <- merge(datshort, justsubjectbert, by="bblid")


#########Total words + word density #########
justwordcount <- cbind(rep(0,31), rep(0,31), rep(0,31), rep(0,31)) %>% as.data.frame()
colnames(justwordcount) <- c("bblid", "numbersentencesall", "meansentlengthall", "wordcount")

for (i in 1:31) {
      temp1 <- filter(datlong, bblid==datshort$bblid[i])
      temp1$bblid <- as.character(temp1$bblid)
      justwordcount[i,1] <- temp1[i,1]
      justwordcount[i,2] <- dim(temp1)[1]
      justwordcount[i,3] <- sum(temp1$s2length)/dim(temp1)[1]
      justwordcount[i,4] <- sum(temp1$s2length)
}

datshort <- merge(datshort, justwordcount, by="bblid")
datshort$worddensity = datshort$wordcount / datshort$recording_duration_sec

##Checking for outliers in the TLC-sum score
boxplot(datshort$tlcsum)$out
datshort[datshort$tlcsum>=15,c(1,70)]
```


## Sample Description
```{r, results="asis"}
tableby(group ~ study + age + sex + race + edu, data = datshort) %>% summary(digits=2, digits.pct=0, digits.p=3)
```
```{r, echo=TRUE}
###Effect sizes for continuous variables
#Age
cohen.d(datshort$age ~ datshort$group)
#Edu
cohen.d(datshort$edu ~ datshort$group)

###Race by group test
fisher.test(datshort$race, datshort$group)
```

## Comparing Cohort 1 vs. Cohort 2
```{r, results="asis"}
tableby(study ~ group + age + sex + race + edu + recording_duration_sec + numbersentencesall + 
              meansentlengthall + wordcount + worddensity, data = datshort) %>% summary(digits=2, digits.pct=0, digits.p=3)
```
```{r}
cohen.d(datshort$recording_duration_sec ~ datshort$study)
```

## Bert + basic language characteristics by subject btw groups
```{r, results="asis"}
tableby(group ~  meanbert + meanweightbert + recording_duration_sec + numbersentencesall + 
              meansentlengthall + wordcount + worddensity, data = datshort) %>% summary(digits=2, digits.pct=0, digits.p=3)
```

### Normality of language measures
- Only mean bert is normal - all others are NOT normally distributed
```{r, echo=TRUE}
shapiro.test(datshort$meanbert)
shapiro.test(datshort$meanweightbert)

shapiro.test(datshort$tlcglobal)
shapiro.test(datshort$tlcsum)
```

### Unweighted BERT
```{r}
lm(meanbert ~ group + study + edu, data = datshort) %>% summary
lm(meanbert ~ group + sex + age + study + race + edu, data = datshort) %>% summary
cohen.d(datshort$meanbert ~ datshort$group)
```

### Weighted within subject
```{r}
lm(meanweightbert ~ edu, data = datshort) %>% summary

lm(meanweightbert ~ group + study, data = datshort) %>% summary
lm(meanweightbert ~ group + study + edu, data = datshort) %>% summary
lm(meanweightbert ~ group + sex + age + study + race + edu, data = datshort) %>% summary
cohen.d(datshort$meanweightbert ~ datshort$group)
```

### Effect size calculations (itemwise for supplemental table)
```{r, echo=TRUE}
for (i in 50:73) {
      print(colnames(datshort)[i])
      cohen.d(datshort[,i] ~ datshort$group) %>% print
      print("----")
}
```

## Calculating slopes for BERT embedding distance by group
```{r}
justBERTdif_SSD <- filter(justBERTdif, Group=="SCZ")
lm(Embedding_Difference ~ Distance_from_Prompt, data=justBERTdif_SSD) %>% summary()
lm(Embedding_Difference ~ Distance_from_Prompt, data=justBERTdif_SSD) %>% confint()

justBERTdif_HC <- filter(justBERTdif, Group=="HC")
lm(Embedding_Difference ~ Distance_from_Prompt, data=justBERTdif_HC) %>% summary()
lm(Embedding_Difference ~ Distance_from_Prompt, data=justBERTdif_HC) %>% confint()

```


## Calculating slopes for BERT embedding distance by group wihtout outliers
```{r}
justBERTdif_rmout_SSD <- filter(justBERTdif_rmout, Group=="SCZ")
lm(Embedding_Difference ~ Distance_from_Prompt, data=justBERTdif_rmout_SSD) %>% summary()
lm(Embedding_Difference ~ Distance_from_Prompt, data=justBERTdif_rmout_SSD) %>% confint()

justBERTdif_rmout_HC <- filter(justBERTdif_rmout, Group=="HC")
lm(Embedding_Difference ~ Distance_from_Prompt, data=justBERTdif_rmout_HC) %>% summary()
lm(Embedding_Difference ~ Distance_from_Prompt, data=justBERTdif_rmout_HC) %>% confint()

```

## Plotting
- summaries of median and interquartile range
```{r}
ggplot(datshort, aes(group, tlcglobal)) + 
      geom_dotplot(aes(fill=group, color=group), binaxis = "y", stackdir="center", alpha=0.4, dotsize=1) +
      stat_summary(fun.y = median, geom="point", aes(color=group), pch=15, size=5) +
      stat_summary(fun.data =median_hilow, mult=1, 
                   geom="errorbar", width=0.2, aes(color=group), lwd=1) +
      theme_bw(base_family = "Times", base_size = 22) +
      scale_color_manual(values=c("lightsteelblue3", "slateblue4"), labels=c("HC", "SSD")) +
      scale_fill_manual(values=c("lightsteelblue3", "slateblue4")) +
      labs(x=NULL, y="TLC Global Score") +
      theme(legend.title = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
      coord_cartesian(ylim = c(-0.51,3.51))

ggplot(datshort, aes(group, tlcsum)) + 
      geom_dotplot(aes(fill=group, color=group), binaxis = "y", stackdir="center", alpha=0.4, dotsize=1) +
      stat_summary(fun.y = median, geom="point", aes(color=group), pch=15, size=5) +
      stat_summary(fun.data = median_hilow, mult=1, 
                   geom="errorbar", width=0.2, aes(color=group), lwd=1) +
      theme_bw(base_family = "Times", base_size = 22) +
      scale_color_manual(values=c("darkolivegreen2", "chartreuse4"), labels=c("HC", "SSD")) +
      scale_fill_manual(values=c("darkolivegreen2", "chartreuse4")) +
      labs(x=NULL, y="TLC Total Score") +
      theme(legend.title = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
      coord_cartesian(ylim = c(-4.6,40.6))

ggplot(datshort, aes(group, meanweightbert)) + 
      geom_dotplot(aes(fill=group, color=group), binaxis = "y", stackdir="center", alpha=0.4, dotsize=1.5) +
      stat_summary(fun.y = median, geom="point", aes(color=group), pch=15, size=5) +
      stat_summary(fun.data = median_hilow, mult=1, 
                   geom="errorbar", width=0.2, aes(color=group), lwd=1) +
      theme_bw(base_family = "Times", base_size = 22) +
      scale_color_manual(values=c("thistle3", "purple4"), labels=c("HC", "SSD")) +
      scale_fill_manual(values=c("thistle3", "purple4")) +
      labs(x=NULL, y="BERT Coherence Score") +
      theme(legend.title = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
      coord_cartesian(ylim = c(.83,1.02))

```

